PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-06-17 manuscript_00.Rmd
new text

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                             TRUE ~ CellType),  
         Tissue = case_when(grepl('Iris', Tissue) ~ 'Iris',
                            TRUE ~ Tissue),
         Tissue = case_when(grepl('^RPE|^Choroid', Tissue) ~ 'RPE-Choroid',
                            TRUE ~ Tissue),
         Citation = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'Swamy VS, Fufa TD, Hufnagel RB, McGaughey DM Building the Mega Single Cell Transcriptome Ocular Meta-Atlas Gigascience 2021 Oct 13;10(10)',
                              TRUE ~ Citation),
         PMID = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ '34651173',
                          TRUE ~ PMID),
         study_accession = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'SRP329495',
                                     TRUE ~ study_accession))

@@

The scater::aggregateAcrossCells function was used to aggregate gene counts across species, study accession, and [category] where category was either Cluster, CellType, or CellType (Predict) assignments for a cell. If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test. Any genes with a sum of 0 counts after aggregation were removed. The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + [category]". Contrasts were extracted either with the target category (e.g. "Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function. Genes were considered to be differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2. To enhance the specificity of the

@@

The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX. There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.

@@

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamy_building_2021]. We updated the scEiaD resource for the PLAE v0.91 web app in six important ways. First, we added a large chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300 (CHECK THIS), to improve the database quality. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons.

@@

web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')

@@

#```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}

#```

## Differential testing identifies thousands genes enriched across various ocular cell types

We used a "pseudo bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping. The creates a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices. This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021]. Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes. We tested each species (Human, Macaque, Mouse) separately. We had an insufficient number of chicken studies to conduct differential testing. The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential expression. After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g. cones versus all remaining cell types) or in a pair-wise manner (cone against rod). We fin
```{r}
# fix  occasional NA
# data made from src/ensdb.R
diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')

@@

diff_tab %>% 
  group_by(Base, Organism) %>% 
  summarise(Count = n()) %>% 
  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
  as_tibble() %>% 
  flextable()

# number of celltype (predict) - organism - study counts (with >=50 cells)
meta_filter %>%  
  filter(Organ == 'Eye') %>% 
  group_by(organism, CellType_predict, study_accession) %>% 
  count() %>% 
  filter(n >= 50) %>% 
  ungroup() %>% 
  group_by(organism, CellType_predict) %>% 
  count() %>% 
  pivot_wider(values_from = n, names_from = organism) %>% 
  filter(!is.na(CellType_predict)) %>% 
  replace(is.na(.), values = 0) %>% 
  rename(`CellType (Predict)` = CellType_predict) %>% 
  flextable()
```

@@

```{r}
scEiaD_2020_v01 %>% 
    tbl("diff_testing") %>% 
    filter(Base == 'Cone') %>%  
    filter(Group == 'CellType (Predict)', 
           Against == 'All') %>% 
  group_by(Gene) %>% 
  summarise(Count = n(), log2FoldChange = mean(log2FoldChange), padj_count = sum(padj < 1e-4), padj = mean(padj)) %>% 
  filter(!is.na(log2FoldChange), padj_count > 1) %>%  
  arrange(-log2FoldChange)

@@

### heatmap human
```{r, fig.width=6, fig.height=17}
# diff_ratios <- diff_tab_big %>% 
#     filter(Against != 'All') %>% 
#     group_by(Symbol, ID, Organism, Base) %>% 
#     summarise(Count = n()) %>% 
#     filter(Count > 1) %>% 
#     left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
#     mutate(Ratio = `Count.x` / `Count.y`) %>% 
#     select(Symbol:Base, Ratio)

gene_sets <- diff_tab %>% 
  #left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
  #filter(Ratio > 0.2) %>% 
  group_by(Base, Organism) %>% 
  slice_max(n=5, order_by = log2FoldChange)

org <- 'Homo sapiens'

g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()

# only uses CT that exist in the organism <-> CellType 
ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()

x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Gene %in% g) %>% 
  collect() %>% 
  filter(Base %in% ct,
         Group == 'CellType (Predict)', 
         Organism == org, Against == 'All') %>% 
  select(Gene, log2FoldChange, Base) %>% 
  pivot_wider(values_from=log2FoldChange, names_from = Base) 
celltype <- colnames(x[,-1])
x <- x %>% data.frame()
row.names(x) <- x$Gene
x <- x[,-1]
Heatmap(x, column_labels = celltype)
```

### heatmap mouse
```{r, fig.width=5, fig.height=20}

org <- 'Mus musculus'

g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()

# only uses CT that exist in the organism <-> CellType 
ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()

x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Gene %in% g) %>% 
  collect() %>% 
  filter(Base %in% ct,
         Group == 'CellType (Predict)', 
         Organism == org, Against == 'All') %>% 
  select(Gene, log2FoldChange, Base) %>% 
  pivot_wider(values_from=log2FoldChange, names_from = Base) 
celltype <- colnames(x[,-1])
x <- x %>% data.frame()
row.names(x) <- x$Gene
x <- x[,-1]
Heatmap(x, column_labels = celltype)
```

### heatmap macaque
```{r, fig.width=5, fig.height=20}

org <- 'Macaca fascicularis'

g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()

# only uses CT that exist in the organism <-> CellType 
ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()

x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Gene %in% g) %>% 
  collect() %>% 
  filter(Group == 'CellType (Predict)', 
         Organism == org, Against == 'All', Base %in% ct) %>% 
  select(Gene, log2FoldChange, Base) %>% 
  pivot_wider(values_from=log2FoldChange, names_from = Base) 
celltype <- colnames(x[,-1])
x <- x %>% data.frame()
row.names(x) <- x$Gene
x <- x[,-1]
Heatmap(x, column_labels = celltype)
```

@@

meta_filter %>% group_by(Citation, PMID, organism, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
```

>>>>

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
# data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
# data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
data_dir_vGiga <- '~/data/scEiaD_gigascience/'
data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                             TRUE ~ CellType),  
         Tissue = case_when(grepl('Iris', Tissue) ~ 'Iris',
                            TRUE ~ Tissue),
         Tissue = case_when(grepl('RPE|Choroid', Tissue) ~ 'RPE-Choroid',
                            TRUE ~ Tissue),
         Citation = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'Swamy VS, Fufa TD, Hufnagel RB, McGaughey DM Building the Mega Single Cell Transcriptome Ocular Meta-Atlas Gigascience 2021 Oct 13;10(10)',
                              TRUE ~ Citation),
         PMID = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ '34651173',
                          TRUE ~ PMID),
         study_accession = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'SRP329495',
                                     TRUE ~ study_accession))

@@

@@

The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX. There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which takes the optimal parameters from SnakePOP to build the scEiaD sqlite database for the app.

@@

Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamy_building_2021]. We updated the scEiaD resource for the PLAE v0.90 web app in six important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`). Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300 (CHECK THIS), to improve the database quality. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons.

@@

tab_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')

@@

```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}

## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types

We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping. The created a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices and thus allowing for mature bulk RNA-seq tooling to be used. This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021]. Once the aggregated matrix was con structured, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes. We tested each species (Human, Macaque, Mouse) separately. We had an insufficient number of chick studies to conduct differential testing. The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression. After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g. cones versus all remaining cell types) or in a pair-wise manner (cone against rod). We find, in total, thousands of differentially expressed genes across the ocular cell types. 
```{r}
# fix  occasional NA
# data made from src/ensdb.R
hs_v86_tab <- read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz')
mm_v79_tab <- read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz')
hs_mm <- bind_rows(hs_v86_tab, mm_v79_tab)

diff_tab_big <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
  filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) > 2) %>% 
  collect() %>% 
  separate(Gene, into = c('Symbol', 'ID'), sep = ' ') %>% 
  mutate(ID = gsub('\\(|\\)', '', ID))

diff_tab_big2 <- diff_tab_big %>% 
  filter(Symbol == '-' | Symbol == 'NA')

diff_tab_big2 <- 
  diff_tab_big2 %>% 
  left_join(hs_mm, by = c('ID' = 'GENEID')) %>% 
  select(-Symbol) %>% 
  rename(Symbol = SYMBOL) %>% 
  relocate(Symbol)

diff_tab_big <- bind_rows(
  diff_tab_big %>% filter(!Symbol %in% c('-', 'NA')),
  diff_tab_big2
)

@@

diff_tab %>% 
  group_by(Base, Organism) %>% 
  summarise(Count = n()) %>% 
  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
  as_tibble() %>% 
  flextable()
```

@@

```{r}
spec_diff <- diff_tab_big %>% 
  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
  group_by(Symbol, Organism, Base) %>% 
  summarise(Count = n()) %>% 
  summarise(Count = n())

spec_diff_go <- spec_diff %>% 
  filter(Count <= 3) %>% 
  mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
  pull(go)

diff_tab_big %>% 
  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
  group_by(Symbol, Organism, Base) %>% 
  summarise(Count = n(),
            baseMean = mean(baseMean),
            log2FoldChange = mean(log2FoldChange),
            padj = mean(padj)) %>% 
  filter(Count > 1) %>% 
  left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
  mutate(Ratio = `Count.x` / `Count.y`) %>% 
  select(-`Count.x`,-`Count.y`) %>% 
  #mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
  #filter(go %in% (spec_diff_go))  %>% 
  #select(-go) %>% 
  ungroup() %>% 
  group_by(Organism, Base) %>% 
  slice_max(n = 10, order_by = log2FoldChange)

@@

### heatmap
```{r, fig.width=5, fig.height=20}
diff_ratios <- diff_tab_big %>% 
    filter(Against != 'All') %>% 
    group_by(Symbol, ID, Organism, Base) %>% 
    summarise(Count = n()) %>% 
    filter(Count > 1) %>% 
    left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
    mutate(Ratio = `Count.x` / `Count.y`) %>% 
    select(Symbol:Base, Ratio)

gene_sets <- diff_tab %>% 
  left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
  filter(Ratio > 0.2) %>% group_by(Base, Organism) %>% 
  slice_max(n=20, order_by = log2FoldChange)

g <- gene_sets %>% filter(Organism == 'Homo sapiens') %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()

x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Homo sapiens', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base) 
celltype <- colnames(x[,-1])
x <- x %>% data.frame()
row.names(x) <- x$Gene
x <- x[,-1]
Heatmap(x, column_labels = celltype)
```

@@

meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
```
